---
title: "Assignment 3: Data Exploration"
author: "Emily Chen"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Exploration
M3: Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing
    `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill
    each instruction.
4.  Assign a useful **name to each code chunk** and include ample
    **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code
    into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the
    dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by
manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()`
or `View()` commands exist in your code.

------------------------------------------------------------------------

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your
    current working directory and upload two datasets: the ECOTOX
    neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and
    the Niwot Ridge NEON dataset for litter and woody debris
    (NEON_NIWO_Litter_massdata_2016-26_raw.csv). Name these datasets
    "Neonics" and "Litter", respectively.

```{r setup,  message = FALSE}
#re-installed all the packages

#install.packages("tidyverse")
#install.packages("lubridate")
#install.packages("here")

#load necessary packages
library(tidyverse)
library(lubridate)
library(here)

#check the current working directory
getwd() # R's current working directory
here() # the project root (where the .Rproj file lives)

#import datasets
Neonics <- read.table(file = here(
  "Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.txt"), header = TRUE, sep = "|", 
  quote = "\"", comment.char = "")
Litter <- read.csv("Data/Raw/NEON_NIWO_Litter_massdata_2016-26_raw.csv")
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental
    Protection Agency's ECOTOX Knowledgebase, a database for
    ecotoxicology research. Neonicotinoids are a class of insecticides
    used widely in agriculture. The dataset that has been pulled
    includes all studies published on insects. Why might we be
    interested in the ecotoxicology of neonicotinoids on insects? Feel
    free to do a brief internet search if you feel you need more
    background information.

> Answer: Neonicotinoids do not only kill target insects but also have a
> much wider impact. First, as insecticides, they expose non-target
> insects, such as bees and other beneficial pollinators to compounds
> that weaken them and reduce their survival. Second, killing insects
> can disrupt entire food webs, as declining insect populations starve
> birds, amphibians, and fish, with effects propagating further up the
> food chain. Third, insecticides can negatively affect natural
> ecosystems more broadly as they can be washed into waterways easily
> and can persist in soils for months, contaminating both soil and
> water. Fourth, overuse drives target pests to evolve resistance,
> rendering the insecticides ineffective over time. We need those data
> to inform the development of more effective insecticides. Finally,
> ecotoxicology research provides the data governments need to justify
> bans and restrictions on insecticides, and to adjust related
> environmental protection laws.

3.  The Niwot Ridge litter and woody debris dataset was collected from
    the National Ecological Observatory Network, which collectively
    includes 81 aquatic and terrestrial sites across 20 ecoclimatic
    domains. 32 of these sites sample forest litter and woody debris,
    and we will focus on the Niwot Ridge long-term ecological research
    (LTER) station in Colorado. Why might we be interested in studying
    litter and woody debris that falls to the ground in forests? Feel
    free to do a brief internet search if you feel you need more
    background information.

> Answer: The study of litter and woody debris is critical because they
> serve as the foundational engine for nutrient cycling, biodiversity,
> and ecosystem resilience. First, the creation and accumulation of
> litter and woody debris allows the forest to form its own nutrient
> cycle. Therefore, measuring litter quantity and chemistry can tell us
> about the health and productivity of a forest's nutrient cycling.
> Second, from the internet research, woody debris creates microhabitats
> for fungi, insects, small mammals, and amphibians. It is often cited
> as a key indicator of forest structural complexity and biodiversity,
> as old-growth forests tend to have much more of it than heavily
> managed or young forests. Third, litter and woody debris can help
> mitigate natural hazards by absorbing the impact of rainfall,
> preventing soil erosion, and reducing water evaporation from the soil.

4.  How is litter and woody debris sampled as part of the NEON network?
    Read the NEON_Litterfall_UserGuide.pdf document to learn more. List
    three pieces of salient information about the sampling methods here:

> Answer: 1.They are respectively collected from elevated and ground
> traps. 2.Locations of tower plots are selected randomly within the 90%
> flux footprint of the primary and secondary airsheds (and addional
> areas in close proximity to the airshed, as necessary to accommodate
> suﬃcient spacing between plots). 3.Ground traps are sampled once per
> year. Target sampling frequency for elevated traps varies by
> vegetation present at the site, with frequent sampling (1x every
> 2weeks) in deciduous forest sites during senescence, and in-frequent
> year-round sampling (1x every 1-2 months) at evergreen sites.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}
dim(Neonics)       # rows, then columns

# 8748 rows and 76 columns
```

6.  Using the `summary` function on the "Effect" column, determine the
    most common effects that are studied. Why might these effects
    specifically be of interest? [Tip: The `sort()` command is useful
    for listing the values in order of magnitude...]

```{r Q6}
# Summary the measurement column 'Effect'
summary(Neonics$Effect) 
#determine the most common effects

sort(table(Neonics$Effect) , decreasing = TRUE) 
#Use table() here as it counts the occurrences of values in categorical variables

```

> Answer: The three most commonly studied effects are mortality,
> population, and genetics. We might be interested in these effects
> because they are among the most straightforward to measure and can be
> quantified using standardized metrics, making it possible to compare
> and contrast results across many studies.

7.  Using the `summary` function, determine the six most commonly
    studied species in the dataset (common name). What do these species
    have in common, and why might they be of interest over other
    insects? Feel free to do a brief internet search for more
    information if needed.[TIP: Explore the help on the `summary()`
    function, in particular the `maxsum` argument...]

```{r Q7}
# Test if the column contains factor values
is.factor(Neonics$X.Species.Common.Name) #return FALSE

# Change the character value to factor
Neonics$X.Species.Common.Name <- as.factor(Neonics$X.Species.Common.Name) 

#Use the summary function to determine the six most commonly studied species
summary(Neonics$X.Species.Common.Name, maxsum = 7)
```

> Answer: The six most commonly studied species in the dataset are Honey
> Bee, Carniolan Honey Bee, Parasitic Wasp, Bumble Bee, Italian
> Honeybee, and Asiatic Honey Bee. Four of the six species are Honey
> Bee. The other is also bee, and one outlier is the Parasitic Wasp.
> While they are different in species, they share in common that they
> are all agriculturally beneficial insects as the bees are pollinators,
> and the wasp is used in biological pest control. Because the interests
> they bring in economic and agricultural, they might be of interest
> over other insects. As the pollination is so important to crop and
> other vegetations, any chemical that threatens their health will be
> studied.

8.  Concentrations are always a numeric value. What is the class of
    `Conc.1.Mean..Author..` column in the dataset, and why is it not
    numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}
#Check the class of the column
class(Neonics$X.Conc.1.Mean..Author..) #Character class
```

> Answer: Instead of numeric, it is a character class. It is not numeric
> highly because of the symbol like '/' is in the data. While unsure why
> slash is in the data, this non-numeric character has turn the entire
> column into character class.

## Explore your data graphically (Neonics)

9.  Create a bar graph of Endpoint counts. What are the two most common
    end points, and how are they defined? Consult the
    ECOTOX_CodeAppendix for more information.

[**TIP**: Add
`theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to
the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}
#Install and read ggplot package
  #install.packages("ggplot2")
library(ggplot2)

#Create a bar plot showing Endpoint counts
ggplot(Neonics,aes(x=Endpoint.)) +
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "Count of Endpoint Categories",
       x = "Endpoint",
       y = "Count")
```

> Answer: The two most common end points are 'LOEL' and 'NOEL'.
> 'LOEL'(Lowest-observable-effect-level) is defined as the lowest dose
> (concentration) producing effects that were significantly different
> (as reported by authors) from responses of controls (LOEAL/LOEC).
> 'NOEL' (No-observable-effect-level) is defined as the highest dose
> (concentration) producing effects not significantly different from
> responses of controls according to author's reported statistical test
> (NOEAL/NOEC).

## Explore your data (Litter)

10. Determine the class of collectDate. Is it a date? If not, change to
    a date and confirm the new class of the variable. Using the `unique`
    function, determine which dates litter was sampled in August 2018.

```{r Q12}
#Check the class of the column
class(Litter$collectDate) #Character class
#Change the Date Class
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
#Confirm the new class of the column
class(Litter$collectDate)
#List the unique value
unique(Litter$collectDate)  #"2018-08-02", "2018-08-30"

#"2018-08-02" and "2018-08-30" are the two dates that                                                 litter was sampled in August 2018.
```

11. Using the `unique` function, list the different plotIDs sampled at
    Niwot Ridge. How is the information obtained from `unique` different
    from that obtained from `summary`?

```{r Q13}
#List the unique value
unique(Litter$plotID) 

```

> Answer: The list of different plotIDs sampled at Niwot Ridge are:
> "NIWO_047" "NIWO_061" "NIWO_062" "NIWO_057" "NIWO_058" "NIWO_064"
> "NIWO_051" "NIWO_067" "NIWO_063" "NIWO_041" "NIWO_046" "NIWO_040"
> 'unique' returns the distinct values themselves and removes duplicate
> elements/rows. While 'summary' returns each distinct value along with
> how many times it occurs.

12. Create a bar graph of functionalGroup counts. This shows you what
    type of litter is collected at the Niwot Ridge sites. Notice that
    litter types are fairly equally distributed across the Niwot Ridge
    sites.

```{r Q14}
#Create a bar plot showing functionalGroup counts
ggplot(Litter,aes(x=functionalGroup)) +
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "Count of Functional Group Types",
       x = "Functional Group Type",
       y = "Count")

```

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: 
Needles

13. Write a short data-exploration summary for Neonics.

> -   **One key characteristic:**
> -   **One data-quality issue:**
> -   **One interesting pattern:**
> -   **One question for further analysis:**

One key characteristic I learned from the Neonics dataset is that the most 
commonly studied species are bees and wasps, which provide significant economic 
and agricultural benefits to humans. One data-quality issue I noticed is that 
the Conc.1.Mean..Author.. column, which should logically be numeric, 
is stored as character data because some entries contain non-numeric 
symbols (the slash symbol), forcing R to read the entire column as text rather 
than as usable numeric values. One interesting pattern is that the most commonly 
studied effects tend to be those that are easier to measure in a standardized 
way. Since bees,especially honey bees, are the most commonly studied insects, 
it would be interesting to explore whether these subspecies show meaningfully 
different sensitivity to neonicotinoids, or whether they react similarly upon 
exposure to the insecticide.

14. Write a short data-exploration summary for Litter.

> -   **One key characteristic:**
> -   **One data-quality issue:**
> -   **One interesting pattern:**
> -   **One question for further analysis:**

One key characteristic I learned from the Litter dataset is that the sampling 
method varies considerably depending on litter type, including the type of trap 
used, the frequency of sampling, and the location of the traps. One data-quality 
issue is that the collection date is stored as a character class rather than 
a date class. One interesting pattern is that litter types appear fairly evenly 
distributed across the Niwot Ridge sites. One question for further study is: 
Given that sampling frequency differs so much by site and vegetation type, 
how might this affect comparisons of litterfall timing or quantity across sites?