---
title: "Assignment 6: Data Scraping"
author: "Emily Chen"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
#load packages
#install.packages('rvest') reinstall the 'rvest' package
library(tidyverse)
library(rvest)
library(here)

#Verify project directory
getwd() # R's current working directory
here() # the project root (where the .Rproj file lives)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
#Fetch the web resources from the URL
webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
webpage
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
#water system name
Water_System_Name <- webpage |>
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") |> 
  html_text()
Water_System_Name
#PWSID
PWSID <- webpage |>
  html_nodes("td tr:nth-child(1) td:nth-child(5)") |> 
  html_text()
PWSID
#Ownership
Ownership <- webpage |>
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") |> 
  html_text()
Ownership
# MGD
MGD <- webpage |>
  html_nodes("th~ td+ td") |> 
  html_text()
MGD
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
#scrape a row
month_withdrawl <- webpage |>
  html_nodes(".fancy-table:nth-child(30) tr+ tr th") |> html_text()
month_withdrawl

df_withdrawals <- data.frame( "Water_System_Name" = rep(Water_System_Name),
                              "PWSID" = rep(PWSID),
                              "Ownership" = rep(Ownership),
                              "Month" = c(1,5,9,2,6,10,3,7,11,4,8,12), 
                              "Year" = rep(2024,12),
                              "Max_Day_Use"= as.numeric(MGD))
df_withdrawals <- df_withdrawals |>
  mutate(Date = my(paste(Month,"-",Year)))

#5 
#create the line plot
ggplot(df_withdrawals,aes(x=Date,y=Max_Day_Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2024 maximum daily withdrawals for",Water_System_Name),
       y="Maximum daily withdrawals (MGD)",
       x="Date") 

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape.it <- function(PWSID, Year) {
#Retrieve the website contents 
the_website <- read_html(
  paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
         PWSID, '&year=', Year))
    #Set the element address variables
  Water_System_Name_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  Ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  MGD_tag <- 'th~ td+ td'
  #Scrape the data items
  Water_System_Name <- the_website |> html_nodes(Water_System_Name_tag) |> 
    html_text()
  Ownership <- the_website |> html_nodes(Ownership_tag) |>  html_text()
  MGD <- the_website |> html_nodes(MGD_tag) |> html_text()
  #Convert to a dataframe
  df_withdrawals <- data.frame( "Water_System_Name" = rep(Water_System_Name, 12),
                                "Ownership" = rep(Ownership, 12),
                               "PWSID" = rep(PWSID, 12),
                               "Month" = c(1,5,9,2,6,10,3,7,11,4,8,12), 
                               "Year" = rep(Year,12),
                               "Max_Day_Use"= as.numeric(MGD)) |> 
    mutate(Date = my(paste(Month,"-",Year)))
  
  #scraping etiquette
  Sys.sleep(1)
  
  #Return the dataframe
  return(df_withdrawals)}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
#Run the function
the_df <- scrape.it('03-32-010',2020)
# make the plot
ggplot(the_df,aes(x=Date,y=Max_Day_Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste(the_df$Year, "maximum daily withdrawals for",
                     the_df$Water_System_Name),
    y="Maximum daily withdrawals (MGD)",
    x="Date")

str(the_df)
```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
#scrape the data for Cary
the_df_cary <- scrape.it('03-92-020',2020)
#bind two dataset
combined_df <- bind_rows(the_df, the_df_cary)
#create plot
ggplot(combined_df, aes(x = Date, y = Max_Day_Use, color = Water_System_Name)) +
  geom_line() +
  geom_point() +
  labs(title = "2020 maximum daily withdrawals: Durham vs. Cary",
       x = "Date",
       y = "Maximum daily withdrawals (MGD)",
       color = "Water system")
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
#create a variable for the years 2018 through 2023
the_years <- 2018:2023
#create another variable that can repeat the PWSID of Cary for six times
the_pwsids <- rep('03-92-020', length(the_years))
#"Map" the "scrape.it" function to retrieve data for all these
#pipe the output of the map2() function to `bind_rows()`
df_cary <- map2(the_pwsids, the_years, scrape.it) |> bind_rows()

#construct plot with tehh smooth line
ggplot(df_cary,aes(x=Date,y=Max_Day_Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste(df_cary$Water_System_Name, 
  "maximum daily withdrawals from 2018 to 2023"),
    y="Maximum daily withdrawals (MGD)",
    x="Date")
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:
> From the plot, it seems like Cary has a strong seasonal pattern where peaks 
occur in the summer and then falls in the winter to the lowest point around 
January or February. From the smoothed line, the overall trend rises gently from 
2018 to 2022, where there is a major rise from 2021 to 2022. After 2022, the line
flattens out and almost stable.
