---
title: "Assignment 2: Coding Basics"
author: "Emily Chen"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Explorations on reproducibility and coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

##Table of Contents

*Add a table of contents*

```{=latex}
\tableofcontents
```

## Crafting Reports - Part 1

### Import packages and data, suppressing messages

Set the following R code chunk so that it runs when knit, but no messages, errors, or any output is shown. The code itself, however, should be displayed.

```{r project.setup, results='hide', warning = FALSE, message = FALSE, echo = TRUE}
#Import libraries
library(tidyverse)
library(knitr)

```


## Coding Basics - Part 2

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 

five_sequence <- seq(from = 1, to = 55, by = 5) #create the sequence

#2. 

mean(five_sequence) #compute the mean of this sequence
median(five_sequence) #compute the median of this sequence

#3. 

five_determine <- if (mean(five_sequence) > median(five_sequence)) 
{'the mean is greater than the median'} else 
    {'the mean is equal or less than the median'} 

#create an if-else statement that compare the mean and the median of the sequence.          Since the question only asks the R to determine whether the mean is greater than             the median, therefore I create the statement to categorize the equal and less                   circumstance together.

```

## Coding Basics - Part 3

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}

student_name <- c("Alice", "Betty", "Cindy", "Daisy") #character data
student_test <- c(90,85,70,88) #numeric data
student_scholarship <- c(TRUE, FALSE, FALSE, TRUE) #Logical Data

Student_records <- data.frame(Name=student_name, Test_Score=student_test, 
                              Scholarship=student_scholarship) 
#create the data frame while label the columns with informative titles
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:

A matrix is a two dimensional structure that can only contain one type of elements. While a data frame is also a two dimensional structure, it can contain multiple types of elements across different columns.

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail". 

11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

12. Run both functions using the value 52.5 as the input

13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#10. Create a function using if...else

test_a <-function(x) {if (x > 50) {"Pass" } else {"Fail"}}

#11. Create a function using ifelse()

test_b <- function(x) {ifelse(x > 50, "Pass", "Fail")}

#12a. Run the first function with the value 52.5
test_a (52.5)
#12b. Run the second function with the value 52.5
test_b (52.5)
#13a. Run the first function with the vector of test scores
  
#test_a(student_test) comments out because this doesn't work

#13b. Run the second function with the vector of test scores
test_b(student_test)
```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")

> Answer:
ifelse() worked. According to the search result on web, 'if'...'else' is not vectorized, which means this function cannot operate on vectors or datasets at once and can only evaluate the first element of the vector. 'ifelse' is vectorized and it can operate element-by-element across an entire vector, returning a vector of the same length.


## Crafting reports - Part 4

### Knit and submit.

Add a table of contents to your document and knit to a PDF. Submit your PDF to Canvas, but also be sure to commit and push your Rmd file used to create this knit document to GitHub. In the section below, add a link to your GitHub repository. 

### Github link

> Answer:
https://github.com/emilychenXC/EDE_R_Lab_F26.git

**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)