---
title: "Assignment 4: Data Wrangling"
author: "Emily Chen"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this assignment, you will build a reproducible workflow that cleans four EPA
air-quality files, combines them, reshapes the result, and produces grouped
summaries. Date conversion is a brief review from Module 3; the emphasis is on
data wrangling.

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.

## Set up your session 

### Questions 1–2

1a. Load the `tidyverse` and `here` packages into your session.

1b. Check your working directory.

1c. Read all four raw files associated with the EPA Air dataset. See the README
file for information about the variables and measurements.

2. Add the appropriate code to reveal the dimensions of the four datasets. 

```{r setup questions 1-2, message = FALSE}
#1a 
library(tidyverse)
library(here)

#1b 
getwd() # R's current working directory
here() # the project root (where the .Rproj file lives)

#1c 
EPAair_O3_NC2024_raw <- read.csv(file=here("Data/Raw/EPAair_O3_NC2024_raw.csv"))
EPAair_O3_NC2025_raw <- read.csv(file=here("Data/Raw/EPAair_O3_NC2025_raw.csv"))
EPAair_PM25_NC2024_raw <- 
  read.csv(file=here("Data/Raw/EPAair_PM25_NC2024_raw.csv"))
EPAair_PM25_NC2025_raw <- 
  read.csv(file=here("Data/Raw/EPAair_PM25_NC2025_raw.csv"))

#2 
#EPAair_O3_NC2024_raw
colnames(EPAair_O3_NC2024_raw)
head(EPAair_O3_NC2024_raw)
summary(EPAair_O3_NC2024_raw)
glimpse(EPAair_O3_NC2024_raw)
dim(EPAair_O3_NC2024_raw)
#EPAair_O3_NC2025_raw
colnames(EPAair_O3_NC2025_raw)
head(EPAair_O3_NC2025_raw)
summary(EPAair_O3_NC2025_raw)
glimpse(EPAair_O3_NC2025_raw)
dim(EPAair_O3_NC2025_raw)
#EPAair_PM25_NC2024_raw
colnames(EPAair_PM25_NC2024_raw)
head(EPAair_PM25_NC2024_raw)
summary(EPAair_PM25_NC2024_raw)
glimpse(EPAair_PM25_NC2024_raw)
dim(EPAair_PM25_NC2024_raw)
#EPAair_PM25_NC2025_raw
colnames(EPAair_PM25_NC2025_raw)
head(EPAair_PM25_NC2025_raw)
summary(EPAair_PM25_NC2025_raw)
glimpse(EPAair_PM25_NC2025_raw)
dim(EPAair_PM25_NC2025_raw)
```

> All four datasets should have the same number of columns (PM2.5 has one extra column),
but unique record counts (rows). Do your datasets follow this pattern?
#Yes

## Wrangle individual datasets to create processed files.

### Questions 3–5

3. Change the Date columns to be date objects.

4. Select the following columns: Date, Daily.AQI.Value, Local.Site.Name, 
AQS.Parameter.Description, County, Site.Latitude, Site.Longitude

5. For the PM2.5 datasets, fill all cells in AQS.Parameter.Description with "PM2.5" 
(all cells in this column should be identical).

```{r questions 3 thru 5}
#3
#EPAair_O3_NC2024_raw
EPAair_O3_NC2024_raw$Date <- mdy(EPAair_O3_NC2024_raw$Date)
#EPAair_O3_NC2025_raw
EPAair_O3_NC2025_raw$Date <- mdy(EPAair_O3_NC2025_raw$Date)
#EPAair_PM25_NC2024_raw
EPAair_PM25_NC2024_raw$Date <- mdy(EPAair_PM25_NC2024_raw$Date)
#EPAair_PM25_NC2025_raw
EPAair_PM25_NC2025_raw$Date <- mdy(EPAair_PM25_NC2025_raw$Date)

#4
#EPAair_O3_NC2024_raw
EPAair_O3_NC2024_raw <- EPAair_O3_NC2024_raw|> 
  select(Date, Daily.AQI.Value, Local.Site.Name, AQS.Parameter.Description, 
         County, Site.Latitude, Site.Longitude)
#EPAair_O3_NC2025_raw
EPAair_O3_NC2025_raw <- EPAair_O3_NC2025_raw|> 
  select(Date, Daily.AQI.Value, Local.Site.Name, AQS.Parameter.Description, 
         County, Site.Latitude, Site.Longitude)
#EPAair_PM25_NC2024_raw
EPAair_PM25_NC2024_raw <- EPAair_PM25_NC2024_raw|> 
  select(Date, Daily.AQI.Value, Local.Site.Name, AQS.Parameter.Description, 
         County, Site.Latitude, Site.Longitude)
#EPAair_PM25_NC2025_raw
EPAair_PM25_NC2025_raw <- EPAair_PM25_NC2025_raw|> 
  select(Date, Daily.AQI.Value, Local.Site.Name, AQS.Parameter.Description, 
         County, Site.Latitude, Site.Longitude)

#5
#EPAair_PM25_NC2024_raw
EPAair_PM25_NC2024_raw$AQS.Parameter.Description <- "PM2.5"
unique(EPAair_PM25_NC2024_raw$AQS.Parameter.Description) #check if PM2.5 fill
#EPAair_PM25_NC2025_raw
EPAair_PM25_NC2025_raw$AQS.Parameter.Description <- "PM2.5"
unique(EPAair_PM25_NC2025_raw$AQS.Parameter.Description) #check if PM2.5 fill
```

### Question 6

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r question 6}
#6 
#EPAair_O3_NC2024
write.csv(
  EPAair_O3_NC2024_raw, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_NC2024_processed.csv")
)
#EPAair_O3_NC2025
write.csv(
  EPAair_O3_NC2025_raw, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_NC2025_processed.csv")
)
#EPAair_PM25_NC2024
write.csv(
  EPAair_PM25_NC2024_raw, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_PM25_NC2024_processed.csv")
)
#EPAair_PM25_NC2025_raw
write.csv(
  EPAair_PM25_NC2025_raw, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_PM25_NC2025_processed.csv")
)

```


## Combine datasets

### Question 7

7. Stack the four datasets with `bind_rows()`. Check that their column names are
identical first.

```{r question 7}
#7 
#check if column names for O3_NC2024_raw and O3_NC2025_raw are identical
identical(colnames(EPAair_O3_NC2024_raw), colnames(EPAair_O3_NC2025_raw)) #TRUE
#check if column names for O3_NC2024_raw and PM25_NC2024_raw are identical
identical(colnames(EPAair_O3_NC2024_raw), colnames(EPAair_PM25_NC2024_raw))#TRUE
#check if column names for O3_NC2024_raw and PM25_NC2025_raw are identical
identical(colnames(EPAair_O3_NC2024_raw), colnames(EPAair_PM25_NC2025_raw))#TRUE

#stack the four datasets
EPAair_combined <- bind_rows(EPAair_O3_NC2024_raw, EPAair_O3_NC2025_raw,
                            EPAair_PM25_NC2024_raw, EPAair_PM25_NC2025_raw)
```

### Question 8

8. Use one pipeline (`%>%`) to create a daily site-level dataset that meets the
following conditions:

  * Include only sites that the four data frames have in common:  
    
  "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue",  
  "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", 
  "West Johnston Co.", "Garinger High School", "Castle Hayne", 
  "Pitt Agri. Center", "Bryson City", "Millbrook School" 
  
  (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Group by date, site name,
  AQS parameter, and county, then calculate the mean AQI value, latitude, and
  longitude. Handle missing values explicitly in each mean.
  
  * Add numeric `Month` and `Year` columns from `Date`. Check if results are integers.
  
  * Hint: the dimensions of this dataset should be 15,012 × 9.

```{r question 8}
#8 
#before setting up pipeline, I decide to create a helper collection of all 
#the sites the question required

sites_collection <- c( "Linville Falls", "Durham Armory", "Leggett", 
"Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", 
  "West Johnston Co.", "Garinger High School", "Castle Hayne", 
  "Pitt Agri. Center", "Bryson City", "Millbrook School" )

Site_level_dataset <- EPAair_combined |> 
  filter(Local.Site.Name %in% sites_collection) |>
  group_by(Date, Local.Site.Name, AQS.Parameter.Description, County) |>
  summarise(Daily.AQI.Value = mean(Daily.AQI.Value, na.rm = TRUE),
    Site.Latitude  = mean(Site.Latitude, na.rm = TRUE),
    Site.Longitude = mean(Site.Longitude, na.rm = TRUE), .groups = "drop") |>
  mutate(
  Month = as.integer(format(Date, "%m")),
  Year  = as.integer(format(Date, "%Y")))

```

### Questions 9–10

9. Use `pivot_wider()` so that ozone and PM2.5 AQI values occupy separate
columns. Each location-date combination should occupy one row.

10. Call up the dimensions of your new tidy dataset.

```{r questions 9-10}
#9
Site_level_dataset.wide <- 
  Site_level_dataset |>
  pivot_wider(
    names_from = AQS.Parameter.Description,
    values_from = Daily.AQI.Value
    )

#10
dim(Site_level_dataset.wide)
summary(Site_level_dataset.wide) #also use summary() to check new columns
```

### Question 11

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC2425_Processed.csv"

```{r question 11}
#11
write.csv(
  Site_level_dataset.wide, 
  row.names = FALSE,
  file = here("Data/Processed/EPAair_O3_PM25_NC2425_Processed.csv")
)

```

## Generate summary tables

### Questions 12–13

12. Create a monthly summary data frame grouped by site, month, and year.
Calculate mean ozone and PM2.5 AQI for each group, handling missing values
explicitly. Then use `drop_na()` to remove rows only when mean **ozone** is
missing. Mean PM2.5 may remain missing.

13. Call up the dimensions of the summary dataset.

```{r questions 12 and 13}
#12
Monthly_EPAair_data_frame <- Site_level_dataset.wide |> 
  group_by(Local.Site.Name, Month, Year) |>
  summarise(PM2.5 = mean(PM2.5, na.rm = TRUE),
    Ozone  = mean(Ozone, na.rm = TRUE),
     .groups = "drop") |> drop_na(Ozone)

#13
dim(Monthly_EPAair_data_frame)
summary(Monthly_EPAair_data_frame) # use summary() to check new data frame
```

### Question 14

14. Why is `drop_na(mean_ozone)` more appropriate here than `na.omit()`? Test
both approaches and compare the dimensions of the resulting summary data frames.

> Answer: 

```{r test}

# in order not to destroy anything, I will create one new data frame to test 
#the na.omit() approach and compare the dimensions

Monthly_EPAair_data_frame_omit <- Site_level_dataset.wide |> 
  group_by(Local.Site.Name, Month, Year) |>
  summarise(PM2.5 = mean(PM2.5, na.rm = TRUE),
    Ozone  = mean(Ozone, na.rm = TRUE),
     .groups = "drop") |> na.omit()

# check the dimensions of the na.omit() data frame through dim() and summary()
dim(Monthly_EPAair_data_frame_omit)
summary(Monthly_EPAair_data_frame_omit)

```

When using drop.na(Ozone), it will only check the Ozone column, and the rows are 
dropped only if Ozone is NA. It will ignore other columns who have NA, such as
PM2.5. Therefore, there are 5 columns/variables, and 238 rows. There are 4 rows
that have PM2.5 data as NA.
However, when using na.omit(), it removes any row with NA in all the columns of 
the data frame. In return, the 4 rows that have PM2.5 data as NA are been 
removed, and there is now only 5 columns and 234 rows. Since the question only
asks for removing rows only when mean ozone is missing and mean PM2.5 may 
remain missing. As a result, in this question, drop_na(Ozone) is more 
appropriate here than na.omit().


