---
title: "Assignment 5: Data Visualization"
author: "Emily Chen"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This assignment accompanies the Environmental Data Analytics module on data
visualization. You will create, format, combine, and interpret plots using
environmental datasets. Your goal is not only to produce correct code, but also
to make deliberate choices about themes, color, accessibility, and the message
communicated by each figure.

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
7. All figures should include clear labels and units, readable text, and color
   choices appropriate for the type of variable displayed. Do not rely on color
   alone when another cue (such as shape, linetype, or facets) would make the
   figure more accessible.
8. Complete your own code before using AI in the designated comparison activity
   in Question 7. Include the prompt and your evaluation of the AI response.

---

## Set up your session 

1. Set up your session. Load the `tidyverse`, `here`, `cowplot`, and any color
palette package(s) you choose to use. Verify your project directory. Read in the
processed NTL-LTER nutrients and chemistry/physics file for Peter and Paul Lakes
(`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`) and the processed
Niwot Ridge litter file (`NEON_NIWO_Litter_mass_trap_Processed.csv`) from the
`Processed_KEY` folder.

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
#load four packages
library(tidyverse)
library(here)
library(cowplot)
library(viridis)

#Verify project directory
getwd() # R's current working directory
here() # the project root (where the .Rproj file lives)

#read files
PeterPaul <- read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"))
NIWO <- read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"))

#2 
#check the format of dates for both dataframe
class(NIWO$collectDate) #returns character
NIWO$collectDate <- ymd(NIWO$collectDate)
class(NIWO$collectDate) #double check the class and returns date

class(PeterPaul$sampledate) #returns character
PeterPaul$sampledate <- ymd(PeterPaul$sampledate)
class(NIWO$collectDate) #double check the class and returns date
```

## Define your theme
3. Build a theme and set it as your default theme. Start with a complete theme,
such as `theme_minimal()` or `theme_classic()`, and then customize at least two
of the following:
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

Choose a base font size that will remain readable when the plots are knitted.
Briefly identify the two or more elements you changed and explain how the
changes improve readability.

```{r build plot theme}
#3
course_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    axis.text = element_text(color = "black")
  )
```

> Theme choices and rationale:
I set the base font size to 12pt and made the plot title bold so that the text 
remains readable and also looks attractive. I removed the minor gridlines to 
reduce visual clutter and let the data stand out. Based on my personal 
preference, I positioned the legend at the right side of the plots and set 
the axis text color to black

## Create graphs
For Questions 4–7, create `ggplot2` graphs and adjust their aesthetics using
data-visualization best practices. Match the color palette to the variable type,
label variables and units clearly, and remove formatting that does not support
the comparison.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), using
separate aesthetics for Peter and Paul lakes. Do not rely only on color: also
use shape, linetype, facets, or another appropriate visual cue. Add a separate
line of best fit for each lake using the `lm` method. Zoom in so that the central
pattern is visible without deleting the extreme observations from the analysis
(hint: use `coord_cartesian()` rather than `xlim()` or `ylim()`). Include a
descriptive title, axis labels with units, and a meaningful legend title.

```{r plot total P vs PO4}
#4 
#set the size and transparency of point
#add a trend line to data and adds a shaded confidence interval
#add cividis color palette
#add graph title and axis titles
#separate two lakes from color and shape of the points
#Zoom in so that the central pattern is visible without 
#deleting the extreme observation (there is one outlier)
#add a legend
#apply the default theme
P_PO43 <- ggplot(PeterPaul, aes(po4, tp_ug,
                                      color = lakename, shape = lakename)) +
  geom_point(alpha = 0.75, size = 2.8) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE) +
  scale_color_viridis_d(option = "viridis") +
  labs(
    title = "Phosphate and phosphorus concentrations vary with lake",
    x = expression("Phosphate"~(mu*g/L)),
    y = expression("Total phosphorus"~(mu*g/L)),
    color = "Lake Name",
    shape = "Lake Name"
  ) + coord_cartesian(xlim = c(0, 100), ylim = c(0, 150)) +
  course_theme

P_PO43

```

Question: What relationship do you observe between phosphate and total
phosphorus? Does the relationship appear to differ between lakes? Identify any
extreme observations and explain why zooming with `coord_cartesian()` is
preferable to removing them with scale limits.

> Answer:
There is a positive relationship between phosphate and total phosphorus. 
As phosphate concentration increases, total phosphorus tends to increase as well. 
This relationship appears similar between the two lakes. 
One extreme outlier is present in Paul Lake. There is an observation with 
a phosphorus concentration within the normal range but an unusually high 
phosphate concentration of 373 mu*g/L. We use coord_cartesian() rather 
than xlim()/ylim() because coord_cartesian() only changes the visible plotting 
window and does not remove any data before the statistics are calculated. 
In contrast, xlim()/ylim() filter the data out entirely before any calculations 
happen, which would alter the fitted trend line and misrepresent the true 
relationship in the data.

5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c)
TN, with month on the x-axis and lake represented by a consistent aesthetic.
Use a qualitative, color-vision-friendly palette. Then create a `cowplot` that
combines the three graphs. Make sure that only one legend is present, the month
order is correct, graph axes are aligned, and each y-axis states the variable
and units.

Tips:
* Month should appear in calendar, not alphabetical, order. You will need to convert 
to factor to order it.
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
#run cowplot library
library(cowplot)
#create boxplot No.1 for month and temperature
T_Month <- PeterPaul |> mutate(month=month(sampledate)) |> 
  mutate(month= factor(month, levels = 1:12)) |>
  ggplot(aes(x = month, y = temperature_C, color = lakename))+
  geom_boxplot() +
  labs(x = 'Month', y = expression("Temperature"~(degree*C))) +
  scale_color_viridis_d(option = "viridis") + course_theme 

#T_Month (I used to use this to test if individual plot works)

#create boxplot No.2 for month and tp
TP_Month <- PeterPaul |> mutate(month=month(sampledate)) |> 
  mutate(month= factor(month, levels = 1:12)) |>
  ggplot(aes(x = month, y = tp_ug, color = lakename))+
  geom_boxplot() +
  labs(x = 'Month', y = expression("Total phosphorus"~(mu*g/L))) +
  scale_color_viridis_d(option = "viridis") + course_theme 

#TP_Month (I used to use this to test if individual plot works)

#create boxplot No.3 for month and tn
TN_Month <- PeterPaul |> mutate(month=month(sampledate)) |> 
  mutate(month= factor(month, levels = 1:12)) |>
  ggplot(aes(x = month, y = tn_ug, color = lakename))+
  geom_boxplot() +
  labs(x = 'Month', y = expression("Total nitrogen"~(mu*g/L))) +
  scale_color_viridis_d(option = "viridis") + course_theme 

#TN_Month (I used to use this to test if individual plot works)

#add a single legend 
T_TN_TP_legend <- get_legend(T_Month + theme(legend.position = "right"))

#stack three individual plots vertically first
combined_plots <- plot_grid(
  T_Month + theme(legend.position = "none", axis.title.x = element_blank()),
  TP_Month + theme(legend.position = "none", axis.title.x = element_blank()),
  TN_Month + theme(legend.position = "none"),
  ncol = 1,
  align = "v")

#turn into cowplot
T_TN_TP <- plot_grid(combined_plots, T_TN_TP_legend,
  ncol = 2,
  rel_widths = c(1, 0.2)) 

T_TN_TP
```

Question: What do you observe about the center, spread, and potential outliers
of each variable across seasons and between lakes? Support your interpretation
with specific features visible in the boxplots.

> Answer: 
The temperature plot shows a clear seasonal pattern of rise from May and reach
to climax during July and August, decline through September and October, 
and drop sharply by November. There is no data been collected from December to
April. Spread is widest in mid-summer, which indicates more variable 
temperatures during peak season, and narrows again by October. There is no
visible outlier in this plot.
The total phosphorus plot shows a more stable pattern from May to September and
there is no data been collected for the rest of the year. 
Peter Lake consistently shows a larger IQR (with a taller box) than Paul Lake,
and its median sits visibly higher than Paul Lake's since June.
Both lakes show extensive high-value outliers every month. Peter lake has a 
higher and larger number of outliers and peak around July with several 
points above 200 mu*g/L. This suggests Peter Lake experiences more frequent 
and more extreme high-phosphorus events than Paul Lake.
The total nitrogen plots shows a similar pattern with the phosphorus plot except
there is a sharp decrease in September. In general, Peter lake shows a larger 
IQR than Paul Lake, but its median only sits visibly higher than Paul Lake's in 
August. Both lakes still show extensive high-value outliers every month but 
number of outliers decreased compared to phosphorus plot. The most extreme 
single point being a Peter Lake observation in August exceeds 3000 mu*g/L.


6. [Niwot Ridge] Subset the litter dataset to the `"Needles"` functional group.
Plot dry mass of needle litter by date and distinguish NLCD classes using a
qualitative color aesthetic. You do not need to rename the land-use classes.
Format the date axis, labels, legend, point size, and transparency so that
overlapping observations remain visible.

7. [Niwot Ridge] First, independently create a second version of the plot with
the three NLCD classes separated into facets rather than distinguished by color.
Keep other formatting choices as consistent as possible so that you can compare
the effect of faceting itself.

```{r Plot litter}
# Unsure about the unit of dry mass therefore I didn't put it on the axis title

#6
Litter_NLCD <- NIWO |> filter(functionalGroup=="Needles") |> 
  ggplot(aes(x=collectDate, y= dryMass, color=nlcdClass)) +
  geom_point(size = 2, alpha = 0.4) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
labs(x = 'Collect Date', y = 'Dry Mass', color = 'NLCD Class',
     title="Needle Litter Dry Mass by Date and NLCD Class") +
  scale_color_viridis_d(option = "viridis") + course_theme 

Litter_NLCD
#7
Litter_NLCD_facet <- NIWO |> filter(functionalGroup=="Needles") |> 
  ggplot(aes(x=collectDate, y= dryMass, color=nlcdClass)) +
  facet_wrap(vars(nlcdClass), ncol = 3) +
  geom_point(size = 2, alpha = 0.4) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
labs(x = 'Collect Date', y = 'Dry Mass', color = 'NLCD Class',
     title = "Needle Litter Dry Mass by Date, Faceted by NLCD Class") + 
  scale_color_viridis_d(option = "viridis") + course_theme 

Litter_NLCD_facet
```

## AI comparison

8. After completing your own faceted plot, provide an AI tool with the Question 7 instructions and the relevant variable names. Ask it to propose `ggplot2` code for the same figure. Do **not** replace your original solution with the AI version.

Paste the exact prompt you used below:

> AI tool used: 
ChatGPT

> AI prompt:
I am working on an assignment in RStudio, can you help me write this code?
I will provide all the variable names you need. Question is :
Subset the litter dataset to the `"Needles"` functional group. 
Plot dry mass of needle litter by date and separate three NLCD classes into 
facets. You do not need to rename the land-use classes.
Format the date axis, labels, legend, point size, and transparency so that
overlapping observations remain visible. The dataset name is NIWO. 
Functional group column is 'functionalGroup'. The date column is 'collectDate', 
the dry mass column is 'dryMass', and NLCD class column is 'nlcdClass'.

Run and inspect the AI-generated code, correcting only errors required to make
it run. If you make corrections, identify them. Then compare the two solutions.

```{r AI solution}
# Paste the AI-generated code here. Clearly comment any corrections you make.
NIWO_needles <- NIWO %>%
  filter(functionalGroup == "Needles")

ggplot(NIWO_needles, aes(x = collectDate, y = dryMass)) +
  geom_point(size = 2, alpha = 0.5) +
  facet_wrap(~ nlcdClass) +
  labs(
    x = "Collection Date",
    y = "Dry Mass",
    title = "Dry Mass of Needle Litter by Collection Date"
  ) +
  scale_x_date(date_labels = "%b %Y") +
  theme_minimal()

```

9. Which plot—your solution or the AI solution—more effectively communicates the data? Evaluate at least three specific choices, such as data filtering, facet structure, date-axis formatting, labels and units, theme, accessibility, or unnecessary code. Identify one useful idea from the AI
solution and one choice you would reject or improve. If the solutions are very similar, explain why both approaches are appropriate.

> AI comparison:
I personally feel my solution is more effective and efficient.
Firstly, the AI-solution seperate the fliter and plot section while I pipe
everything together into a clear structure.
Secondly, I do some internet search and find out (~ nlcdClass) is an older version
compared to vars(nlcdClass). Though it has been stated that they are 
functionally equivalent, the newer version can make faceting more consistent
with group_by(), select(), etc. Those are all common functions that we use, so
my solution should be more effective.
Thirdly, I decide to add colors to each NLCD class though it is not asked, but
from the output plot, I think it is more aesthetically preferable.
One thing I really want to learn and improve after seeing AI solution is that
the solution not only contains all the elements the question asked, but also
the whole structure is very clean and straightforward that allow people easily
understand and potentially replicate. However, mine is pretty bulky. 
Also, I suspect because I set legend at the right side of the graph,
my x-axis have overlapped and crushed together.

10. Question: Which encoding in Questions 6 and 7—color or facets—do you think is more effective for comparing NLCD classes, and why? Consider overlap, ease of comparison, accessibility, and the amount of space used by each figure.

> Answer: 
I think color is a more effective way of comparing NLCD classes because it uses 
less space than faceting with the same date breaks, making it easier for the 
audience to read. Although I added color to my own facet plot, the original 
facet plot should look like the AI-generated solution above, which doesn't rely 
on color perception at all. This would decrease accessibility for readers. 
Although the majority of points from each NLCD class fall and overlap in a 
similar area/range, it is still relatively easy to identify and make comparisons 
between each class. Therefore, I think comparing classes by color is more 
effective than faceting in this circumstance.
